---
title: "Introduction to R"
author: "Amanda Zacharias"
output:
  html_document:
    rmdformats::html_clean: 
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This page provides an overview of R syntax. Try to code along!

>For beginners: The help tab in the bottom right corner of RStudio is your best friend! Use it often

## Basic math
```{r}
1 + 2 # addition
1 - 2 # subtraction
1 * 2 # multiplication
1 / 2 # divition
1^2 # exponentiation
exp(2) # exponentiation (natural)
log(2) # logarithm (natural)
log2(2) # logarithm (base 2)
log10(2) # logarithm (base 10)
log(exp(2)) # log() and exp() cancel each other out
abs(-2) # absolute value
```

## Rounding
```{r}
round(1.1387, digits = 2) # round up
floor(1.1387) # lower integer
ceiling(1.1387) # higher integer
signif(1.1387, digits = 2) # significant digits
```

## Logic
```{r}
1 > 2 # greater
1 < 2 # less 
1 >= 2 # greater or equal
1 <= 2 # less or equal
1 == 2 # equal
1 != 2 # not equal
TRUE | FALSE # OR
FALSE | FALSE # OR 
TRUE & FALSE # AND
TRUE & TRUE # AND
```

## Assigning and viewing variables 
Generally `<-` is used to assign variables and `=` is used to assign parameters in functions.
```{r}
# Assign
a <- 1
b <- 2
c <- a + b

# View
a # directly type the variable
print(a) # print to console, and return invisibly (not great inside functions)
paste(a) # convert to string, can be used to write to console but not often
cat(a) # concatenate and print; no fancy formatting but my preferred approach

```

## Combining elements 
```{r}
# Vectors
x <- c("a", "b", "c", "d")
x
y = c("a", "c", 2, "e")
y
```

```{r}
# Lists
lst <- list(x, y)
lst
namedLst <- list("x" = x, "y" = y)
namedLst
```

```{r}
# Dataframes
df <- data.frame("x" = x, 
                 "y" = y)
df
```

## Indexing by numbers
```{r}
# Vectors and lists
x[0] # R is 1-based!
x[1]
x[-1]
x[length(x) - 1]
namedLst[1]
namedLst[[1]]
# Dataframes
df[1, ]
df[1:2, ]
df[, 1]
```

## Indexing by names
```{r}
# String 
namedLst["x"]
namedLst[["x"]]
df["x"]
# $
namedLst$x
df$x
# Range of numbers ======
1:10
1:length(x)
seq(1 ,10 , 2)
```

## Random 
```{r}
set.seed(0)
runif(5)
sample(1:10, 1)
```

## Is x in y? 
```{r}
x %in% y # literally "is x in y?"
match(x, y) # where is x first in y ? 
match(c("a", "c"), c("a", "a", "b"))
```

## If else 
```{r}
num <- 10
if (num < 10){
  print("lesser")
} else if (num > 10){
  print("greater")
} else if (num == 10){
  print('equal')
}
```

## Functions 
```{r}
# Simple function
HelloWorld <- function(arg1, arg2="world"){
  #' This function prints arg1 and arg2 to the console
  #' 
  #' @param arg1 First string
  #' @param arg2 Second string
  #' @return Prints a message to the console
  #' @example
  #' HelloWorld("Hello")
  cat(arg1, arg2)
  return(arg1)
}
HelloWorld("Hello")

# Addition
Add <- function(num1, num2){
  #' This function adds two numbers together
  #'
  #' @param num1 A number
  #' @param num2 A second number
  #' @return Returns the sum of the two numbers
  sum <- num1 + num2
  return(sum)
}
Add(1, 2)
```

## Loops
```{r}
# For 
for (obj in x){
  print(obj)
}
for (idx in 1:length(x)){
  print(idx)
  print(x[idx])
}
```

```{r}
# While
idx <- 1
while (idx <= length(x)){
  print(idx)
  print(x[idx])
  idx = idx + 1
}
```

## Working directory
```{r}
getwd()
#setwd()
```

## Input/Output
```{r}
# Common input functions
csv <- read.csv("foo.csv")
table <- read.table("foo.txt", header = TRUE)

# Common output functions
write.csv(csv, "practice.csv")
write.table(table, "practice.txt")
```

## Install packages
```{r}
# Install CRAN
# install.packages("dplyr")

# Install Bioconductor
# if (!require("BiocManager", quietly = TRUE)){
#   install.packages("BiocManager")
# }
# BiocManager::install("dplyr")
```

```{r message=FALSE}
# Load
library("dplyr") # version 1.1.2
# Inspect version
packageVersion("dplyr")
```






































